A Critical Review of Sophistication and Algorithmic Model Selection

### 2014-12-17 Facticity/Facticity/paper.tex
gezeur van steven

@@

\begin{abstract}
Kolmogorov complexity is a sound expression of the information content of a binary string, but it leaves open the issue of \emph{meaningful information}. The strings with maximal information content are not the ones we would regard as interesting, or meaningful. One attempt to capture the idea of meaningful information is \emph{sophistication}. It is a simple idea with many attempted formalizations: describe the string efficiently with a two-part coding approach, and take the complexity of the model as a measure of the amount of meaningful information. We describe three requirements that any formalization must meet in order to satisfy the expectations that come with the idea of sophistication: invariance, unboundedness and efficiency. We show that no currently published proposal meets these requirements \pb{TODO}, and conjecture that no formalization of the idea of sophistication can hope to do so. We conclude by reviewing the consequences of this conjecture.
\end{abstract}
\section{Introduction}

\noindent Kolmogorov complexity provides an objective measurement of the information content of an object. Any arbitrary choices made in its construction can be changed, without significantly changing the outcome. This property is crucial for the measure's success: it allows us to see Kolmogorov complexity as a \emph{property of the data} rather than an arbitary function computed on it. This, in turn, allows us to define the intrinsic entropy, or randomness, inherent in a single object expressed as a binary string, without any assumptions on the process that produced it. Unfortunately, this idea of complexity does not fit one-on-one to our intuitive idea of which objects are complex.

Consider the following, illustrative example: an old-fashioned, analog television set which can be tuned to three frequencies. The first produces a blank screen, the second a regular program, and the third white noise. By the Kolmogorov complexity, these  broadcasts have increasing information content. The first is easiest to describe, the second contains a great deal of regularity, and the third cannot be compressed. However, in terms of meaningful information, information that we find interesting, the second signal is by far the richest. This notion has proved far harder to capture in formal terms than the simple notion of `raw' information content.

One of the most frequently explored ideas for capturing the notion of meaningful information is \emph{sophistication}. The idea behind sophistication is to select a description for the data by two-part coding (describing a model first, and then the data in terms of the model), and to take the complexity of only the model part as a measure of the meaningful information. The assumption is that both simple and random data have small models; in the case of the simple data, a small model describes all regularity, and no further information is required, and in the case of the random data, the data itself is its shortest representation, and the model is a `print function' which simply reproduces its input. If the dataset is more interesting, however, containing both regularity and randomness, its best representation will have a more complex model. A key assumption of the idea of sophistication is that we can perform objective model selection for a given string. That is: there is a function that, given a string, selects a single unique model that can be used to compress this string optimally. Moreover, any elements of this function that are chosen ad hoc can be changed without significantly affecting the outcome.

In the case of the Kolmogorov complexity, this holds true; we can change the universal Turing machine used in its definition and the complexities returned will not change by more than a constant. This is known as \emph{invariance}. Because Kolmogorov complexity is invariant, we can view it as an intrinsic property of the data, rather than simply an arbitrary function computed on it. If a formalization of sophistication is ever to be held in the same regard as the Kolmogorov complexity, it must be subject to a similar principle of invariance.

The unifying principle over all the notions we review in this paper is simple. We start with an object expressed as a finite binary string. We represent this object, hoping to compress it, in two parts, the first is a model, and the second an `input' to this model. How the two parts are encoded, and what the class of models is are details that differ between variants. We know, however, that for successful representations, those that capture all structure in the string, the length of the two-part code should approach the Kolmogorov complexity. We take the set of all \emph{sufficient statistics}---the models of those representations that are `close enough' to the Kolmogorov complexity---and call the one with the lowest complexity the \emph{minimal sufficient statistic}. The complexity of the minimal sufficient statistic is the sophistication.

The intuition is that both very simple and very random data have small, simple programs. The strings with much structure and many interesting properties are those for which the best model is large. Two assumptions are crucial to this view:
\begin{itemize}
  \item The two-part coding efficiently separates the structure and the noise into distinct parts. In the optimal representation the model contains only `structural information' and the noise contains only the `uninteresting' information.
  \item This principle is invariant. If we change, for instance, the universal Turing machine used to unpack our descriptions, the minimal sufficient statistic will not change significantly.
\end{itemize}

In addition to

In this article we will show that most proposals do not achieve these goal. For one remaining case, we can offer no such proof, but there exists no proof to the contrary either. We offer several insights that suggest that the idea of an objective sophistication function is an unattainable ideal.

@@

We will first briefly discuss the different variants of sophistication in a historical context, to give the reader a simple, chronological overview, without discussing too much technical detail. In the main body of the paper, we will define each variant rigorously, and in non-chronological order.

The first suggestion that a representation of a string can be separated cleanly into a structural part, containing a description of its regularities, and a noise part, containing the residual information, came from Kolmogorov himself, in a talk given at a conference in Tallinn in 1973 (first committed to paper by Cover \cite{todo} and later more extensively discussed by Vitanyi \cite{todo}). Here, he proposed what has become known as the \emph{Kolmogorov structure function}. Briefly, this function operates on a space of two part representations for a given string $x$, consisting of some finite set $S$ (containing $x$) and the index of $x$ in $S$. The \emph{minimal sufficient statistic}, it was proposed, is the set with the lowest complexity for which the length of the two-part representation reaches the prefix-free Kolmogorov complexity. \pb{TODO: lowest randomness deficienty.}

Koppel \cite{} built on this notion, coining the phrase \emph{sophistication}. He changed the class of `models' from finite sets to total functions and used Levin's monotone complexity, instead of prefix-free Kolmogorov complexity. One interesting difference is that Koppel takes the description of the data under a variant of Kolmogorov complexity, which consists of a description of a model and that model's input, and takes the size of the model part as the sophistication. In contrast, Kolmogorov's minimal sufficient statistic is built on a second, two-part, representation in addition to the Kolmogorov complexity's one-part representation. This difference in perspective is subtle but relevant to the discussion ahead.

While Koppel expands the model class from finite sets to functions, he maintains a restriction that the functions be total. This is a recurring theme: if the model class is expanded to all total functions, the the universal Turing machine itself becomes a model, so that we always have a two-part description with a small model, to within a constant of the Kolmogorov complexity. In most treatment this `collapse' to the universal model is avoided by excluding from the model class, using some restriction. This acts as a kind of force, pushing information into the model part of the two-part code. As we shall see, defining sophistication quickly becomes a balancing act, trying to match two forces which by themselves push information one way or the other.

In 1996 Gell-Mann and Lloyd introduced `Effective complexity'\cite{todo}. Working from the perspective of statistical mechanics, they arrived at a notion remarkably similar to Sophistication. Effective complexity describes abinary string in terms of a probability distribution over finite strings (an `ensemble' in the terminology of Gell-Mann and Lloyd) and the (Shannon) entropy of the distribution. The reasoning is that, if the string is typical for the distribution, the entropy is sufficient to describe the string given the distribution, so that the length of the description of the distribution and the size of its entropy suffices as a two-part description. The two-part description which determines the effective complexity is chosen by placing a time bound (taken from the singleton description) on the computation of the string from the distribution. The hope is that this will move deep (ie. hard to compute) information to the model part, and shallow information to the data. In a later treatment\cite{}, Gell-Mann and Lloyd select the decisive two-part representation from those achieving minimal code-length based on ``the average quantities judged to be important'', which would seem to make the value highly subjective and context-dependant.

G\'acs, Tromp and Vit\'anyi \cite{}, built on the idea of the structure function to establish a framework of \emph{algorithmic statistics}, in the hope of founding a statistical theory not on probabilistic notions, but on combinatorial ones. This echoes Kolmogorov's own hope for the theory of information:
\begin{quotation}
\noindent Information theory must precede probability theory and not be based on it. By the very essence of this discipline, the foundations of information theory must have a finite combinatorial character
\end{quotation}
Algorithmic statistics studies binary strings and their possible two-part representations. In a later paper\cite{gacs2001algorithmic}, Vit\'anyi specifically studies the role of the minimal sufficient statistic as an expression of the `meaningful' information of a string (taking the name sophistication from Koppel). Like Koppel, Vit\'anyi restricts the model class to total functions, eliminating the universal model. He balances this force, by allowing the total two-part description to be non-self delimiting, so that both the representation with all information in the model part, and the representation with everything in the residual part are less efficient than the best two-part coding.

>>>>

% in introductie?
%
% model, model class
% two-part code: sophistication is length of first part
% kolmogorov complexity?
% KC = two part vs KC in two part
% collapse, restrictions on model class (eg total instead of partial recursive, see section XX)
% conjecture

@@

\begin{abstract}
\sdr{Emphasize ongoing attempts to formalize meaningful information, ``recurring dream''. Also in the introduction} \sdr{Don't single out sophistication}
Kolmogorov complexity is a sound expression of the information content of a binary string, but it leaves open the issue of \emph{meaningful information}. The strings with maximal information content are not the ones we would regard as interesting, or meaningful. One attempt to capture the idea of meaningful information is \emph{sophistication}. It is a simple idea with many attempted formalizations: describe the string efficiently with a two-part coding approach, and take the complexity of the model as a measure of the amount of meaningful information. We describe three requirements that any formalization must meet in order to satisfy the expectations that come with the idea of sophistication: invariance, unboundedness and efficiency. We show that no currently published proposal meets these requirements \pb{TODO}, and conjecture that no formalization of the idea of sophistication can hope to do so. We conclude by reviewing the consequences of this conjecture.
\end{abstract}

\sdr{Get conjecture right}

\section{Introduction}

\noindent Kolmogorov complexity is an objective measure of the information content of an object. Although its definition contains several arbitrary choices, these only affect its value up to a constant additive term. This property is crucial for the measure's success: it allows us to view Kolmogorov complexity as a \emph{property of the data} rather than an arbitary function computed on it. This, in turn, allows us to define the intrinsic entropy, or randomness, inherent in a single object expressed as a binary string, without any assumptions on the process that produced it. However, this notion of complexity does not in all respects match our intuitive idea of which objects are complex, and which objects are not.

Imagine an old-fashioned, analog television set which can be tuned to three frequencies. The first produces a blank screen, the second an ordinary TV show, and the third white noise. Judging by their Kolmogorov complexity, these  broadcasts have increasing information content. The first is easiest to describe, the second is harder but still contains a great deal of regularity, and the third cannot be compressed. However, intuitively speaking the second signal clearly conveys more meaningful information than the other two. This notion has proved far harder to capture in formal terms than the simple notion of `raw' information content. \sdr{Emphasize recurring dream here}

One of the most frequently explored ideas for capturing the notion of meaningful information is \emph{sophistication}. The idea behind sophistication is to select a description for the data by two-part coding (describing a model first, and then the data in terms of the model), and to take the complexity of only the model part as a measure of the meaningful information. The assumption is that both simple and random data have small models: \sdr{the intuition is that small models suffice to describe both simple and random data, but not the meaningful information we are attempting to quantify.} in the case of the simple data, a small model describes all regularity, and no further information is required, and in the case of the random data, the data itself is its shortest representation, and the model is a `print function' which simply reproduces its input. If the dataset is more interesting, however, containing both regularity and randomness, its best representation will have a more complex model. \sdr{slightly repetitive} A key assumption of the idea of sophistication is that we can perform objective model selection for a given string. That is: there is a function that, given a string, selects a single unique model that can be used to compress this string optimally. Moreover, any elements of this function that are chosen ad hoc can be changed without significantly affecting the size of the preferred model.

In the case of the Kolmogorov complexity, this holds true; we can change the universal Turing machine used in its definition and the complexities returned will not change by more than a constant. This is known as \emph{invariance}. Because Kolmogorov complexity is invariant, we can view it as an intrinsic property of the data, rather than simply an arbitrary function computed on it. If a formalization of sophistication is ever to be held in the same regard as the Kolmogorov complexity, it must be subject to a similar principle of invariance. \sdr{I feel it's not about regard}

The unifying principle over all the notions we review in this paper is simple. We start with an object expressed as a finite binary string. We represent this object, hoping to compress it, in two parts, the first is a model, and the second an `input' to this model. How the two parts are encoded, and what the class of models is are details that differ between variants. We know, however, that for successful representations, those that capture all structure in the string, the length of the two-part code should approach the Kolmogorov complexity. We take the set of all \emph{sufficient statistics}---the models of those representations that are `close enough' to the Kolmogorov complexity---and call the one with the lowest complexity the \emph{minimal sufficient statistic}. The complexity of the minimal sufficient statistic is the sophistication. \sdr{But I thought in some of our settings we use a different baseline than KC!}

The intuition is that both very simple and very random data have small, simple programs. The strings with much structure and many interesting properties are those for which the best model is large.\sdr{Repetitive} Two assumptions are crucial to this view: \sdr{This bit should be more solid}
\begin{itemize}
  \item The two-part coding efficiently \sdr{what does efficiently mean?} separates the structure \sdr{undefined} and the noise into distinct parts. In the optimal representation the model contains only `structural information' and the noise contains only the `uninteresting' information.
  \item This principle is invariant. If we change, for instance, the universal Turing machine used to unpack our descriptions, the minimal sufficient statistic will not change significantly.
\end{itemize}

In this article we will show that most proposals do not achieve these goals. For one remaining case, we can offer no such proof, but there exists no proof to the contrary either. \sdr{Say this more strongly: we believe that ``fixes'' of the problems with the approach are obfuscation rather than real solutions}  We offer several insights that suggest \sdr{insights that suggest stuff sounds very weak} that the idea of an objective sophistication function is an unattainable ideal.

@@

We will first briefly discuss the different variants of sophistication in a historical context, to give the reader a simple, chronological overview, without discussing too much technical detail. \sdr{We first provide a chronological overview of the variants of sophistication and discuss their relationships.} In the main body of the paper, we will define each variant rigorously, and in non-chronological order.

The first suggestion that a representation of a string can be separated cleanly into a structural part, containing a description of its regularities, and a noise part, containing the residual information, came from Kolmogorov himself, in a talk given at a conference in Tallinn in 1973 (first committed to paper by Cover \cite{todo} and later more extensively discussed by Vit\'anyi \cite{todo}). Here, he proposed what has become known as the \emph{Kolmogorov structure function}. Briefly, this function operates on a space of two part representations for a given string $x$, consisting of some finite set $S$ (containing $x$) and the index of $x$ in $S$. The \emph{minimal sufficient statistic} is the set with the lowest complexity for which the length of the two-part representation reaches the prefix-free Kolmogorov complexity. \pb{TODO: lowest randomness deficienty.}

Koppel \cite{} built on this notion, coining the phrase \emph{sophistication}. He changed the class of `models' from finite sets to total functions and used Levin's monotone complexity, instead of prefix-free Kolmogorov complexity \sdr{why?}. One interesting difference is that Koppel takes the description of the data under a variant of Kolmogorov complexity, which consists of a description of a model and that model's input, and takes the size of the model part as the sophistication. In contrast, Kolmogorov's minimal sufficient statistic is built on a second, two-part, representation in addition to the Kolmogorov complexity's one-part representation. This difference in perspective is subtle but relevant to the discussion ahead.\sdr{Good point but currently much too hard to understand}

\sdr{Following bit is currently pretty vague. I think the collapse idea deserves a better introduction} While Koppel expands the model class from finite sets to functions, he maintains a restriction that the functions be total. This is a recurring theme: if the model class is expanded to all total functions, the the universal Turing machine itself becomes a model, so that we always have a two-part description with a small model, to within a constant of the Kolmogorov complexity. In most treatment this `collapse' to the universal model is avoided by excluding it from the model class, using some restriction. This acts as a kind of force, pushing information into the model part of the two-part code. As we shall see, defining sophistication quickly becomes a balancing act, trying to match two forces which by themselves push information one way or the other.

In 1996 Gell-Mann and Lloyd introduced `Effective complexity'\cite{todo}. Working from the perspective of statistical mechanics, they arrived at a notion remarkably similar to Sophistication. Effective complexity describes abinary string in terms of a probability distribution over finite strings (an `ensemble' in the terminology of Gell-Mann and Lloyd) \sdr{I really think the ensemble is just the set} and the (Shannon) entropy of the distribution. The reasoning is that, if the string is typical for the distribution, the entropy is sufficient to describe the string given the distribution, so that the length of the description of the distribution and the size of its entropy suffices as a two-part description. The two-part description which determines the effective complexity is chosen by placing a time bound (taken from the singleton description) on the computation of the string from the distribution. The hope is that this will move deep (ie. hard to compute) information to the model part, and shallow information to the data. In a later treatment\cite{}, Gell-Mann and Lloyd select the decisive two-part representation from those achieving minimal code-length based on ``the average quantities judged to be important'', which would seem to make the value highly subjective and context-dependent.

G\'acs, Tromp and Vit\'anyi \cite{}, built on the idea of the structure function to establish a framework of \emph{algorithmic statistics}, as an attempt to found a statistical theory not on probabilistic notions, but on combinatorial ones. This echoes Kolmogorov's own hope for the theory of information:
\begin{quotation}
\noindent Information theory must precede probability theory and not be based on it. By the very essence of this discipline, the foundations of information theory must have a finite combinatorial character \sdr{Source?}
\end{quotation}
Algorithmic statistics studies binary strings and their possible two-part representations. In a later paper\cite{gacs2001algorithmic}, Vit\'anyi specifically studies the role of the minimal sufficient statistic as an expression of the `meaningful' information of a string (taking the name sophistication from Koppel). Like Koppel, Vit\'anyi restricts the model class to total functions, eliminating the universal model. He balances this force, by allowing the total two-part description to be non-self delimiting, so that both the representation with all information in the model part, and the representation with everything in the residual part are less efficient than the best two-part coding.\sdr{This balancing act is not yet clear to the reader}
